Group Theory in Physics (Course Note)

### 2016-10-31 Lecture note - Group Theory in Physics/group theory in physics.tex
New lecture note of group theory in physics.

\usepackage{float}  % For H in figures
\usepackage{amsmath} % For math
\usepackage{amssymb}
\usepackage{mathrsfs}
% Followings are for the special character: differential "d".
\newcommand*\diff{\mathop{}\!\mathrm{d}}
\newcommand*\Diff[1]{\mathop{}\!\mathrm{d^#1}}
\numberwithin{equation}{subsection} % have the enumeration go to the subsection level.
                                    % See:https://en.wikibooks.org/wiki/LaTeX/Advanced_Mathematics
\usepackage{graphicx}   % need for figures
\usepackage{cite} % need for bibligraphy.
\usepackage[unicode]{hyperref}  % make every cite a link
\usepackage{CJKutf8} % For Chinese characters
\usepackage{fancyref} % For easy adding figure,equation etc in reference. Use \fref or \Fref instead of \ref
\usepackage{braket} %http://tex.stackexchange.com/questions/214728/braket-notation-in-latex

% Following is for theorems etc environments
% http://tex.stackexchange.com/questions/45817/theorem-definition-lemma-problem-numbering && https://en.wikibooks.org/wiki/LaTeX/Theorems
\usepackage{amsthm}
\newtheorem{defi}{Definition}[section]
\newtheorem{thm}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\newtheorem{remark}{Remark}[section]
\newtheorem{prop}{Proposition}[section]
\newtheorem{coro}{Corollary}[section]
\theoremstyle{definition}
\newtheorem{ex}{Example}[section]
\newtheorem{fact}{Fact}[section]

@@

\section{Anchor}
\label{sec:Anchor}
\begin{thebibliography}{1}
    %\bibitem{book} 
    \bibitem{book} Zhongqi Ma, Group Theory in Physics
    \bibitem{math.se_1_lenz_vector} \href{physics.stackexchange.com/questions/18088/what-symmetry-causes-the-runge-lenz-vector-to-be-conserved}{What symmetry causes the Runge-Lenz vector to be conserved?}
\end{thebibliography}
\printnomenclature
\section{License}
The entire content of this work (including the source code
for TeX files and the generated PDF documents) by 
Hongxiang Chen (nicknamed we.taper, or just Taper) is
licensed under a 
\href{http://creativecommons.org/licenses/by-nc-sa/4.0/}{Creative 
Commons Attribution-NonCommercial-ShareAlike 4.0 International 
License}. Permissions beyond the scope of this 
license may be available at \url{mailto:we.taper[at]gmail[dot]com}.
\end{document}

>>>>

\usepackage{float}  % For H in figures
\usepackage{amsmath} % For math
\usepackage{amssymb}
\usepackage{mathrsfs}
% Followings are for the special character: differential "d".
\newcommand*\diff{\mathop{}\!\mathrm{d}}
\newcommand*\Diff[1]{\mathop{}\!\mathrm{d^#1}}
\numberwithin{equation}{subsection} % have the enumeration go to the subsection level.
                                    % See:https://en.wikibooks.org/wiki/LaTeX/Advanced_Mathematics
\usepackage{graphicx}   % need for figures
\usepackage{cite} % need for bibligraphy.
\usepackage[unicode]{hyperref}  % make every cite a link
\usepackage{CJKutf8} % For Chinese characters
\usepackage{fancyref} % For easy adding figure,equation etc in reference. Use \fref or \Fref instead of \ref

% Following is for theorems etc environments
% http://tex.stackexchange.com/questions/45817/theorem-definition-lemma-problem-numbering && https://en.wikibooks.org/wiki/LaTeX/Theorems
\usepackage{amsthm}
\newtheorem{thm}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\newtheorem{remark}{Remark}[section]
\newtheorem{prop}{Proposition}[section]
\newtheorem{coro}{Corollary}[section]
\theoremstyle{definition}
\newtheorem{defi}{Definition}[section]
\newtheorem{ex}{Example}[section]
\newtheorem{fact}{Fact}[section]

@@

\section{Skipped Two lectures}
Due to GRE physics preparation. Concepts that may have been covered:

\begin{itemize}
    \item Representation of Groups.
        \begin{itemize}
            \item Character of representation.
            \item Equivalence between representation.
        \end{itemize}
    \item Transformation of Fields
    \item Regular Representation.
\end{itemize}

Now let's me cover these concepts quickly.

\begin{defi}[Representation]
\nomenclature{Representation}{\nomrefpage.}
    A representation of a group $G$ is a continuous homomorphism $D$
    from $G$ to the group of automorphisms of a vector space $V$:
    \begin{equation}
        D: G\mapsto \mathrm{Aut}V
    \end{equation}
    $V$ is called the \textit{represetation space}, and the
    \textit{dimension of the representation} is the dimension of $V$.
\end{defi}

The following is copied from \cite{Ludeling}.

\begin{itemize}
    \item There is always the representation $D(g) = 1$ for all $g$.
        If $\mathrm{dim} V = 1$, this is called the \textit{trivial
        representation} \nomenclature{trivial
        representation}{\nomrefpage}.
    \item The matrix groups, i.e. $\mathrm{GL}(n, K)$ and subgroups,
        naturally have the representation "by themselves", i.e.
        by$n\times n$ matrices acting on $K_n$ and satisfying the
        defining constraints (e.g. nonzero determinant). This is
        loosely called the \textit{fundamental or defining
        representation}.
    \item Two representations $D$ and $D'$ are called
        \textit{equivalent}\nomenclature{equivalent of
        representation}{\nomrefpage} if they are related by a
        similarity transformation, i.e. if there is an operator S such
        that
        \begin{equation}
            SD(g)S^{-1} = D'(g)
        \end{equation}
        for all $g$. Note that $S$ does not depend on $g$! Two equivalent
        representation can be thought of as the same representation in
        different bases. We will normally regard equivalent
        representations as being equal.

        Note also here $S$ is the transformation of basis. If we know
        the transformation of vectors $X$, then
        \begin{equation}
            X^{-1} D(g)X = D'(g)
        \end{equation}
    \item A representation is called
        \textit{faithful}\nomenclature{faithful
        representation}{\nomrefpage}if it is injective, i.e.
        $\mathrm{ker} D = \{e\}$, or in other words, if $D(g_1) \neq
        D(g_2)$ whenever $g_1 \neq g_2$.
    \item If $V$ is equipped with a (positive definite) scalar
        product, D is unitary if it preserves that scalar product, i.e
        if
        \begin{equation}
            \braket{u,v} = \braket{D(g)u,D(g)v}
        \end{equation}
        for all $g\in G$. (Here we assume that $V$ is a complex vector
        space, as that is the most relevant case. Otherwise one could
        define orthogonal representations etc.)
\end{itemize}

\section{20161031}
\label{sec:20161031}

\begin{thm}[Unitary Representation]
    For finite groups and for compact Lie groups, all representations
    are equivalent to a unitary representation.
\end{thm}
\begin{proof}
    For $D(g)$, we need to find a $X$ such that $\bar{D}(g)\equiv
    X^{-1}D(g)X$ is unitary.

    Since
    \begin{align*}
        1 = \bar{D}^\dagger \bar{D}
    \end{align*}
    One will find 
    $$ (XX^\dagger)^{-1} = D^\dagger (XX^\dagger)^{-1} D$$
    Then let
    \begin{equation}
        H\equiv \sum_{s\in G} D^\dagger(s) D(s)
    \end{equation}
    One can verify that
    \begin{equation}
        D^\dagger (g) H D(g) = H
    \end{equation}

    Now we construct $X$ from $H$. We have
    \begin{equation}
        H = (XX^\dagger)^{-1}
    \end{equation}

    Notice we have $H$ is Hermitian by above equation. Also, $H$ is
    positive definite (easily seen from the definition of $H$ and 
    $a^\dagger H a \geq 0$. Also $H$ is of full rank.).

    Then we have $UHU^{-1} = \mathrm{diag}\{\gamma_1,\gamma_2,\cdots\}$
    and $\gamma_i > 0$. The rest for constructing $X$ should be
    obvious.

    Another way to prove this is to imbe
\end{proof}

\begin{thm}
    For any two equivalent representation, there is always a unitary
    matrix to relate the two, i.e. exits $Y$ unitary, s.t.
    $$ \bar{D}(g) = Y^{-1} D(g) Y$$
\end{thm}
\begin{proof}
    Suppose we have two unitary representation $D(g)$ and
    $\bar{D}(g)$, related by
    $$ \bar{D}(g) = X^{-1} D(g) X$$
    where $X$ is not necessarily unitary.

    Let $H\equiv X^\dagger X$, it is easy to prove that $H$ is
    Hermitian and positive definite. Direct calculation shows that
    $$ \bar{D}^{-1}(g) H \bar{D}(g) = H$$
    That is $\bar{D}(g)$ and $H$ commute.
    Then we construct $Y$ from $H$. Let $V$ be such that
    \begin{equation}
        V^{-1} H V = \Gamma
    \end{equation}
    where $\Gamma$ is a diagonal matrix of $H$'s eigenvalues.
    Obviously $V$ has to be unitary. Define $Y$ to be
    \begin{equation}
        Y\equiv V \sqrt{\Gamma} V^{-1}
    \end{equation}
    By direct calculation, we have
    $$ (XY)^\dagger (XY) = 1$$
    One can show that
    $$[ Y, \bar{D}(g)] = 0$$
    with laborious calculation. % TODO 
\end{proof}

\begin{defi}[Reducibility of Representation]
    \nomenclature{Reducibility of Representation}{\nomrefpage.}
    A representation $D$ is called \textit{reducible} if $V$ contains an
    invariant subspace. Otherwise $D$ is called irreducible.

    A representation is called \textit{fully reducible} if $V$ can be
    written as the direct sum of irreducible invariant subspaces, i.e.
    $V=V_1\oplus \cdots \oplus V_p$, all the $V_i$ are invariant and the
    restriction of $D$ to each $V_i$ is irreducible.
\end{defi}

\begin{ex}$ $

    \begin{itemize}
        \item The representation of finite group is obviously fully
            reducible or irreducible.
        \item The representation of translation group is not fully
            reducible. For example, for translation we have:
            $T_a T_b = T_{a+b}$,
            One can confirm that the following representation obeys
            the above relationship:
            $$T_a = \left( \begin{array}{cc}
                 1 & a \\
                 0 & 1 \\
            \end{array} \right)$$
            but this is obviously not fully reducible.
    \end{itemize}
\end{ex}

\begin{defi}[Interwiner]
\nomenclature{Interwiner}{\nomrefpage.}
    Given two representations $D_1$ and $D_2$ acting on $V_1$ and
    $V_2$, an intertwiner between $D_1$ and $D_2$ is a linear operator
    \begin{equation}
        F: V_1 \mapsto V_2
    \end{equation}
    which "commutes" with $G$ in the sense that
    \begin{equation}
        F D_1(g) = D_2(g) F
    \end{equation}
    for all $g\in G$.
\end{defi}

(From pp.30 of \cite{Ludeling})

The existence of an intertwiner has a number of consequences. First,
$D_1$ and $D_2$ are equivalent exactly if there exists an invertible
intertwiner. Second, \hl{the kernel and the image of $F$ are invariant
subspaces}: Assume $v\in \Ker{F}$, i.e. $Fv = 0$. Then
\begin{equation}
    FD_1 v = D_2 F v = D_2 0 = 0
\end{equation}
so $D_1 v\in \Ker{F}$. On the other hand, let $w_2 = Fw_1$ be an
arbitrary element of the image of $F$. Then from the definition we have
\begin{equation}
    D_2 w_2 = D_2 F w_1 = FD_1 w_1
\end{equation}
which is again in the image of $F$. Now if $D_1$ is irreducible, the only invariant subspaces,
hence the only possible kernels, are $\{0\}$ and $V_1$ itself, so $F$ is either injective or zero.
Similarly, if $D_2$ is irreducible, F is either surjective or zero. Taking these statements together, we arrive at Schur’s Lemma:

\begin{lemma}[Schur's lemma 1]
    An intertwiner between two irreducible representations is either
    an isomorphism, in which case the representations are equivalent,
    or zero
\end{lemma}

An important special case is the one where $D_1 = D_2$. In that case,
we see that $F$ is essentially unique. More precisely, we have the
following theorem, also often called Schur’s Lemma:
\begin{lemma}[Schur's lemma 2]
    If $D$ is an irreducible finite-dimensional representation on a complex
vector space and there is an endomorphism $F$ of $V$ which satisfies 
\begin{equation}
    FD(g) = D(g) F
\end{equation}
for all $g\in G$, then F is a multiple of the identity, $F = \lambda
\id$
\end{lemma}
\begin{proof}
    Note that $F$ has at least one eigenvector $v$ with eigenvalue
    $\lambda$. (This is where we need $V$ to be a complex vector
    space: A real matrix might have complex eigenvalues, and hence no
    real eigenvectors.) Clearly, $F-\lambda\id$ is also an
    intertwiner, and it is not an isomorphism since it annihilates
    $v$. Hence, by Schur’s Lemma, it vanishes, thus $F = \lambda\id$.
\end{proof}

(From the book \cite{book})
\begin{thm}[Orthogonal Theorem]
    For finite group $G$, let $D^i(G)$ and $D^j(G)$ be its two
    irreducible, non-equivalent, and unitary representation. Then, as
    a vector in group algebra, they have the following orthogonal
    relationship:
    \begin{equation}
        \sum_{R\in G} D^{i*}_{\mu\rho}(R)D^j_{\nu \lambda}(R) =
        \frac{g}{m_j} \delta_{ij} \delta_{\mu\nu}\delta_{\rho\lambda}
    \end{equation}
    $g$ is the order of the group, and $m_j$ is the
    dimension of representation $D^j(G)$.
\end{thm}
\begin{proof}
    Let
    \begin{equation}
        Y_{\rho\lambda}^{\mu\nu} \equiv
        \delta_{\rho\lambda}\delta_{\mu\nu}
    \end{equation}
    Then let
    \begin{align*}
        X^{\mu\nu} = \sum_{R\in G} D^i(R^{-1}) Y^{\mu\nu} D^j(R)
    \end{align*}
    One can find by direct calculation
    \begin{align}
        X^{\mu\nu}_{\rho\lambda}= \sum_{R\in G}
        D^{i*}_{\mu\rho}(R)D^j_{\nu \lambda}(R)
    \end{align}
    And also through direct calculation, one finds
    \begin{align*}
        D^i (s) X^{\mu\nu} = X^{\mu\nu} D^j(s)
    \end{align*}
    for any $s\in G$. 
    Then $X^{\mu\nu}$ is a interwiner. So the case for $i\neq j$ is
    obvious. When $i=j$, we have:
    $$ X = \lambda \id$$
    Now we find the $\lambda$, i.e. the eigenvalue of $X^{\mu\nu}$.

    Now since
    \begin{align*}
        X^{\mu\nu}_{\rho\lambda} = \lambda^{\mu\nu}
        \delta_{\rho\lambda}
    \end{align*}
    One can find two fact by direct calculation:
    \begin{align*}
        &\sum_\rho X^{\mu\nu}_{\rho\lambda} = m_j \lambda^{\mu\nu} \\
        &\sum_\rho X^{\mu\nu}_{\rho\lambda} = g \delta^{\mu\nu}
    \end{align*}
    Hence $\lambda^{\mu\nu} = \frac{g}{m_j} \delta^{\mu\nu}$.
\end{proof}
\begin{remark}
    For the first orthogonality $\delta_{ij}$, one actually do not
    required that $D^i$ and $D^j$ is unitary.
\end{remark}
\begin{coro}
    \begin{equation}
        \sum_{j=1}^{l} m_j^2 \leq g
    \end{equation}
    % TODO
\end{coro}
\begin{coro}
    \begin{equation}
        \sum_{R\in G} \chi^{i*}(R)\chi^j(R) = g \delta_{ij}
    \end{equation}
\end{coro}
\section{Anchor}
\label{sec:Anchor}
\begin{thebibliography}{1}
    \bibitem{book} Zhongqi Ma, Group Theory in Physics
    \bibitem{math.se_1_lenz_vector} \href{physics.stackexchange.com/questions/18088/what-symmetry-causes-the-runge-lenz-vector-to-be-conserved}{What symmetry causes the Runge-Lenz vector to be conserved?}
    \bibitem{Ludeling} Lecture Notes for physics751: Group Theory (for
    Physicists), by C Ludeling.
    \href{http://www.th.physik.uni-bonn.de/nilles/people/luedeling/grouptheory/data/grouptheorynotes.pdf}{Link}
\end{thebibliography}
\printnomenclature
\section{License}
The entire content of this work (including the source code
for TeX files and the generated PDF documents) by 
Hongxiang Chen (nicknamed we.taper, or just Taper) is
licensed under a 
\href{http://creativecommons.org/licenses/by-nc-sa/4.0/}{Creative 
Commons Attribution-NonCommercial-ShareAlike 4.0 International 
License}. Permissions beyond the scope of this 
license may be available at \url{mailto:we.taper[at]gmail[dot]com}.
\end{document}
